prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-09 inst/MyArticle/paper.Rmd
Adding DEqMS for further references. Otherwise integrate CP's comments.

Protein quantification proteomics experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment versus control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs with increasing numbers of samples are studied, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a specific factors and their interactions are significant. Furthermore it can be tested if differences of protein abundance between groups are significant. For this application it is important that the model provides an unbiased estimate of the false discovery rate (FDR) so that protein lists can be selected for follow-up experiments [TODO reference].   
Furthermore, the quantified proteins can be ranked using the estimated fold-change, $t$-statistics or scaled $p$-value, and then subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins.

@@

bb <- data.frame(dplyr::bind_rows(xx))
rownames(bb) <- names(xx)
knitr::kable(data.frame(bb), caption = "Rows - R package, Columns - types of models supported: pd - parallel design, rm - repeated measurements, fd - factorial design, int - interactions among factors, mem - mixed effect models, eb - empirical Bayes, md - missing data modelling. Y - yes, y - yes but only for random effects, * - repeated measurements are only modeled using a fixed effects. ? - the hurdle model is not implemented in the msqrob2 package yet.")
```

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enable us to easily interface with many data manipulation, visualization, and modeling methods implemented in base __R__[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref{fig:tabCompletion}, and easy to use (OOP).

@@

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. As a result, we obtain different fold changes by comparing the various dilutions, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected fold change is 1.2 for E. coli proteins and 1 for human proteins.
We processed the raw data using MaxQuant [cite] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by Maxquant are available in the _prolfquadata_ R package [@prolfquadata].

@@

@@

Given a linear model contrasts can be computed by 
\begin{eqnarray} 
\hat{\beta_{c}} &=& \sum l\beta_{m} 
\end{eqnarray}

and

@@

with $X$ being the design matrix, $\beta_m$ the model parameters and $l$ an array of coefficients. The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model.

For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `r BiocStyle::CRANpkg("lmerTest")` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are implemented in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

@@

A relevant parameter is the number of contracts a given method estimates from the data(see Figure \@ref(fig:FDRfdp) left Panel). It indicates how robust the procedure works in the presence of missing data. For each protein, we can estimate four contrast which for 4046 proteins, at least two peptides resulted in $16184$ possible contrast estimates. Because the set of contrasts slightly differs for each method, it makes comparing them complicated. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ while the mixed effect models estimate the fewest with $15940$. Hence, to conclude that one method shows a better performance, the score, e.g., pAUC, needs to be greater, but also the number of estimated contrasts needs to be equal or larger.

Figure \@ref(fig:FDRfdp) left panel, shows how various estimates obtained from the models, i.e. $\log_2(FC)$, $t$-statistics and scaled $p$-values allow to identify true positives (TP) given a false positive rate. When testing for differential expression, the $\log_2(FC)$ is outperformed by than the $t$-statistic or $p$-value. 
If an accurate estimate of the $log_2(FC)$ is essential, the linear models fitted to protein intensities, estimated using Tukey's median polish, show the best performance. We speculate that this is because of the robustness of Tukey's median polish method. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

There are only minor differences in the $p$-AUC when using the $t$-statistics or the scaled $p$-value  (see Figure \@ref(fig:FDRfdp) right panel). In order to compute the confusion matrices based on the $p$-value we first need to rescale it (see Equation \@ref(eq:scalepvalue)). Thus, the $p$-value is a function of the $t$-statistics and the degrees of freedom. Interestingly, the pAUC increases when using $p$-values instead of the $t$-statistics for linear models, while it decreases for mixed effect models. The reason is erroneous denominator degrees of freedom estimation for many proteins in the mixed-effect models.

We also benchmark, if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:FDRfdp) panel C draws on the horizontal axis the FDR determined from the model, and on the vertical axis, the FDP obtained from the confusion matrix. We see that most of the lines are below the diagonal, indicating that the FDR estimates obtained with `r BiocStyle::Githubpkg("fgcz/prolfqua")` are modestly conservative for this benchmark dataset provide. In the case of `r BiocStyle::Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of `r BiocStyle::Biocpkg("PECA")`, the FDR estimates, obtained when Benjamini-Hochberg correcting the regulation probabilities, strongly overestimate the $FDP$.

@@

We focused in our benchmark on the comparison of the statistical modelling methods and fixing the pre-processing steps. However, there are other equally or even more important parameters of any protein quantification pipeline. One of them is the normalization of the intensities within the samples to remove systematic differences and to make them comparable [@pursiheimo2015optimization]. The method used to infer protein intensities from peptide intensities is an additional important factor [cite TopN]. For instance, the original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ inensities reported by MaxQuant, _proDA_ performs significantly worse ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 73%$). Last but not least, the software [@Cox2008MaxQuant,@yu2020fast], identifying proteins, and generating the quantification values can also significantly contribute to the performance of the entire pipeline, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis.

@@

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A) for the largest number of proteins (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated fold changes are unbiased (not shown) and show high diagnostic accuracy (\@ref(fig:FDRfdp) Panel A). Furthermore, the scaled $p$-values or the $t$-statistics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

(ref:proDAAllBenchmarks) Comparison of fold change estimates produced by proDA and using the missing data modelling method implemented in prolfqua.

>>>>

Protein quantification proteomics experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment versus control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs with increasing numbers of samples are studied, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a specific factors and their interactions are significant. Furthermore it can be tested if differences of protein abundance between groups are significant. For this application it is important that the model provides an unbiased estimate of the false discovery rate (FDR) so that protein lists can be selected for follow-up experiments.   
Furthermore, the quantified proteins can be ranked using the estimated fold-change, $t$-statistics or scaled $p$-value, and then subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins.

@@

bb <- data.frame(dplyr::bind_rows(xx))
rownames(bb) <- names(xx)
knitr::kable(data.frame(bb), caption = "Rows - R package, Columns - types of models supported: pd - parallel design, rm - repeated measurements, fd - factorial design, int - interactions among factors, mem - mixed effect models, eb - empirical Bayes, md - missing data modelling. Y - yes, * - repeated measurements are modeled using a fixed effects. ? - the hurdle model was published but unclear if available in the msqrob2 package.")
```

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enable us to easily interface with many data manipulation, visualization, and modeling methods implemented in base __R__[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref(fig:tabCompletion), and easy to use (OOP).

@@

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. As a result, we obtain different fold changes by comparing the various dilutions, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected fold change is 1.2 for E. coli proteins and 1 for human proteins.
We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by MaxQuant are available in the _prolfquadata_ R package [@prolfquadata].

@@

### Robust scaling of the data

[TODO : Describe robust scaling]

@@

To allow for the comparison of different treatments, given a linear model $y = \beta X$, a contrast is computed by a linear combination $l$ of linear model parameters $\beta_m$. The difference estimate $\beta_c$, is given by:

\begin{eqnarray} 
\hat{\beta_{c}} &=& \sum l\beta_{m} 
\end{eqnarray}

and the variance of $\beta_c$ by:

@@

with $X$ being the design matrix, $\beta_m$ the model parameters and $l$ an array of weights, defining the contrast $c$. The degrees of freedom for the contrast are equal to the residual degrees of freedom of the linear model. For estimating contrasts from mixed effects models we used the function `contest` implemented in the R package `r BiocStyle::CRANpkg("lmerTest")` [@Kuznetsova2017lmerTest] and used the Satterthwaite method to estimate the denominator degrees of freedom. These methods are implemented in the class `Contrast` (see Figure \@ref(fig:ContrastUML))

### Determining linear parameter combinations for treatment comparison

[TODO : Describe how we produce linear combinations from model specification]

@@

A relevant parameter is the number of contracts a given method estimates from the data (see Figure \@ref(fig:FDRfdp) Panel B). It indicates how robust the procedure works in the presence of missing data. For each protein, we can estimate four contrast which for 4046 proteins, at least two peptides resulted in $16184$ possible contrast estimates. Because the set of contrasts slightly differs for each method, it makes comparing them complicated. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ while the mixed effect models estimate the fewest with $15940$. Hence, to conclude that one method shows a better performance, the score, e.g., pAUC, needs to be greater, but also the number of estimated contrasts needs to be equal or larger.

Figure \@ref(fig:FDRfdp) Panel A, shows how various estimates obtained from the models, i.e. the difference between groups, $t$-statistics and scaled $p$-values allow to identify true positives (TP) given a false positive rate. When ordering the proteins by the probablity of differential expression the difference is outperformed by than the $t$-statistic or $p$-value. If an accurate estimate of the difference is essential, the linear models fitted to protein intensities, estimated using Tukey's median polish, show the best performance. We speculate that this is because of the robustness of Tukey's median polish method. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

There are only minor differences in the $p$-AUC when using the $t$-statistics or the scaled $p$-value  (see Figure \@ref(fig:FDRfdp) Panel A). In order to compute the confusion matrices based on the $p$-value we first need to rescale it (see Equation \@ref(eq:scalepvalue)). Thus, the $p$-value is a function of the $t$-statistics and the degrees of freedom. Interestingly, the pAUC increases when using $p$-values instead of the $t$-statistics for linear models, while it decreases for mixed effect models. The reason is erroneous denominator degrees of freedom estimation for many proteins in the mixed-effect models.

We also benchmark, if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:FDRfdp) Panel C draws on the horizontal axis the FDR determined from the model, and on the vertical axis, the FDP obtained from the confusion matrix. We see that most of the lines are below the diagonal, indicating that the FDR estimates obtained with `r BiocStyle::Githubpkg("fgcz/prolfqua")` are modestly conservative for this benchmark dataset provide. In the case of `r BiocStyle::Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of `r BiocStyle::Biocpkg("PECA")`, the FDR estimates, obtained when Benjamini-Hochberg correcting the regulation probabilities, strongly overestimate the $FDP$.

@@

We focused in our benchmark on the comparison of the statistical modelling methods and fixing the pre-processing steps. However, there are other equally or even more important parameters of any protein quantification pipeline. One of them is the normalization of the intensities within the samples to remove systematic differences and to make them comparable [@pursiheimo2015optimization]. The method used to infer protein intensities from peptide intensities is an additional important factor [@Grossmann2010]. For instance, the original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, _proDA_ performs significantly worse ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 73%$). Last but not least, the software [@Cox2008MaxQuant,@yu2020fast], identifying proteins, and generating the quantification values can also significantly contribute to the performance of the entire pipeline, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis.

@@

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A) for the largest number of proteins (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated fold changes are unbiased (not shown) and show high diagnostic accuracy (\@ref(fig:FDRfdp) Panel A). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the linear model with variance moderation. Because of the robustness of the dropout model to missing observations, we intend to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

(ref:proDAAllBenchmarks) Comparison of fold change estimates produced by proDA and produced by linear models implemented in prolfqua.
